Drug sensitivity prediction with normal inverse Gaussian shrinkage informed by external data

### 2020-04-06 docs/supplement1.Rnw

  \section{Estimation of alternative prior}\label{sec:altprior}
  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
  (\ref{md-eq:altprior}) are estimated by empirical Bayes maximisation of
  the marginal log likelihood:
  \begin{subequations}\label{eq:altmll}
    \begin{align}
      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
      \right| \\
      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
      \left\{ \I + \mathbf{X} 
      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
    \end{align}
  \end{subequations}
  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
  \mathbf{c}_{pd} \end{bmatrix} \tr$.
  The first term in the sum of (\ref{eq:altmll}) is concave in 
  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
  the marginal log likelihood a sum of non-concave (as well as non-convex) 
  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
  are generally not well suited to maximise non-concave functions. Maximisation
  is therefore done through a minorise-maximisation (MM) algorithm, 
  similar to \cite{zeng_incorporating_2020}.

@@

  \bibliographystyle{author_short3}  
	\bibliography{refs}

>>>>

@@

  \section{Alternative prior}\label{sec:altprior}
  Consider the following simpler hyperprior alternative to 
  equation (\ref{md-eq:prior}) in the MD:
  \begin{subequations}\label{eq:altprior}
		\begin{align}
		  \beta_{jd} | \gamma^2_{jd}, \tau_d^2, \sigma^2_d & \sim \mathcal{N}_{p} 
		  (0, \gamma_{jd}^2 \tau_d^2 \sigma_d^2), \\
		  \log \gamma_{jd}^{2} & = \mathbf{c}_{jd} \tr \bm{\alpha}_{\text{feat}}, \\
		  \log \tau_d^{2} & = \mathbf{z}_{d} \tr \bm{\alpha}_{\text{drug}}\\
		  \sigma_d^{2} & \sim \Gamma^{-1} (\nu, \zeta).
		\end{align}
	\end{subequations}
	This model is a Bayesian, multivariate extension of 
	\cite{zeng_incorporating_2020}. The model does 
	not propagate uncertainty in the variance components $\gamma_{jd}^{2}$
	and $\tau_d^{2}$, but directly models the variance components as an
	exponentiated linear function of the external covariates. The linear functions
	are modelled on the log-scale to ensure non-negative variance components.
	Additionally, it allows for more flexibility in the error variance
	prior model. Note that in the limit $\zeta \to 0$ and $\nu=1/2$, the error
	variance follows the Jeffrey`s prior from (\ref{md-eq:prior}) in the MD. 
	Throughout the
	following, whenever $\zeta$ and $\nu$ are not specified, we use $\zeta \to 0$ 
	and $\nu=1/2$.

  The hyperparameters $\bm{\alpha} = \begin{bmatrix} \bm{\alpha}_{\text{feat}}
  \tr & \bm{\alpha}_{\text{drug}} \tr \end{bmatrix} \tr$ in model 
  (\ref{eq:altprior}) are estimated by empirical Bayes maximisation of
  the marginal log likelihood:
  \begin{subequations}\label{eq:altmll}
    \begin{align}
      \log p (\bm{\alpha} | \mathbf{Y}) & = \sum_{d=1}^D
      \log \[ \int_{\bm{\beta}_d} \int_{\sigma^2_d} \mathcal{L} 
      (\mathbf{y}_d | \bm{\beta}_d, \sigma^2_d ) \pi(\bm{\beta}_d | \sigma^2_d) 
      \pi(\sigma^2_d) d\bm{\beta}_d d\sigma_d^2 \] \\
      & \propto - \sum_{d=1}^D \bigg[ \log \left|\I + \mathbf{X} 
      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} \tr 
      \right| \\
      & \,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\,\, + (n + 1) \log \( \mathbf{y}_d \tr 
      \left\{ \I + \mathbf{X} 
      \diag \[ \exp\( \tilde{\mathbf{C}}_d \bm{\alpha} \) \] \mathbf{X} 
      \tr\right\}^{-1} \mathbf{y}_d \) \bigg],
    \end{align}
  \end{subequations}
  where $\tilde{\mathbf{C}}_d = \begin{bmatrix} \mathbf{C}_d & 
  \mathbf{1}_{p \times 1} \cdot \mathbf{z}_d \tr \end{bmatrix}$, and
  $\mathbf{C}_d = \begin{bmatrix} \mathbf{c}_{1d} & \cdots & 
  \mathbf{c}_{pd} \end{bmatrix} \tr$.
  The first term in the sum of (\ref{eq:altmll}) is concave in 
  $\bm{\alpha}$, while the second term is convex in $\bm{\alpha}$. This renders
  the marginal log likelihood a sum of non-concave (as well as non-convex) 
  functions. General purpose optimisers, such as \texttt{optim} in \texttt{R} 
  are generally not well suited to maximise non-concave functions. Maximisation
  is therefore done through a minorise-maximisation (MM) algorithm, 
  similar to \cite{zeng_incorporating_2020}.

  \section{Extra simulation results}\label{sec:simulations}
  The lasso and ridge models were fit separately per drug, using the \texttt{R}
	package \texttt{glmnet} \cite[]{friedman_regularization_2010} with cross
	validated
	penalty parameters. The \texttt{xtune} model is fit with the corresponding 
	\texttt{R} package \cite[]{ni_bayesian_2019}.

	Performance measures are estimated from the remaining 914 cell lines

<<simulation_gdsc_res1>>=
res <- read.table("results/simulations_gdsc_res1.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)

tab <- cbind(
  apply(
    aggregate(res[substr(rownames(res), 1, 6)=="emsel.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 6)=="emsel.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 6)=="emseh.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 6)=="emseh.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 5)=="emse.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 5)=="emse.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 5)=="pmse.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 5)=="pmse.", ])),
              FUN="mean")[, -1], 2, median))

tab <- round(tab, 3)
tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
                   "$\\text{EMSE}_{\\text{top}}$",
                   "$\\text{EMSE}_{\\text{total}}$",
                   "$\\text{PMSE}$")     
rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}}^-$",
                            "NIG$_{\\text{f}}$")
@

  $\lambda_{\text{feat}}$ is overestimated with median estimate 
	\Sexpr{round(est.lambdaf[2], 2)} (the
	true value is 1). The overestimated $\lambda_{\text{feat}}$ translates 
	into underestimated $\gamma_{jd}^2$ variances.

	Table \ref{tab:simulations_gdsc_tab1} displays median estimation MSE 
	(EMSE) and prediction MSE (PMSE) for simulation Scenario 1, calculated on an 
	independent test data set. EMSE is further split into the contribution of the
	bottom 90\% of the $\beta_{jd}$ in terms of size (in absolute value) and the 
	top 10\% of the $\beta_{jd}$ in size. The table includes as a reference the 
	null model, which is an empty model that only fits the means of the $\y_{d}$.
	Estimation MSE (EMSE) and prediction MSE (PMSE) give measures of predictive 
	performance. Focussing on MSE, we see that the NIG$_{\text{f}}$ model, 
	that includes the external covariates, performs best, both in terms of 
	estimation and prediction. The lower MSEs confirm that the model that
	includes the external data (NIG$_{\text{f}}$) learns the underlying structure 
	in the data better than model NIG$_{\text{f}}^-$ that does not include the
	external data. Furthermore, we see that estimation of the larger $\beta_{jd}$ 
	is where ridge loses the most compared to the NIG and lasso models. 
	The heavier tails of the NIG and lasso priors allow for larger $\beta_{jd}$
	estimates and thus lower EMSE for these $\beta_{jd}$.

<<simulations_gdsc_tab1>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure for simulation Scenario 1, estimated on the test data (lowest in bold).", 
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<simulation_gdsc_res2>>=
res <- read.table("results/simulations_gdsc_res2.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.chi <- median(1/res[rownames(res)=="alphad0", 1])
est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
tab <- cbind(
  apply(
    aggregate(res[substr(rownames(res), 1, 6)=="emsel.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 6)=="emsel.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 6)=="emseh.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 6)=="emseh.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 5)=="emse.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 5)=="emse.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 5)=="pmse.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 5)=="pmse.", ])),
              FUN="mean")[, -1], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <- 
  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <- 
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <- 
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <- 
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
                   "$\\text{EMSE}_{\\text{top}}$",
                   "$\\text{EMSE}_{\\text{total}}$",
                   "$\\text{PMSE}$")      
rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{d}}^-$",
                            "NIG$_{\\text{d}}$")
@

  $\lambda_{\text{feat}}$ is overestimated with median estimate
	\Sexpr{round(est.lambdaf[1], 2)} (the true value is one) and thus 
	again underestimates the 
	prior $\gamma_{jd}^2$ variance.

	Median $\lambda_{\text{drug}}=\Sexpr{round(est.lambdad[2], 2)}$ is
	overestimated compared to true value one. 
	Scale $\lambda_{\text{drug}}$ is underestimated with median estimate
	\Sexpr{round(est.lambdad[1], 2)} compared to the true value one.

  Median performance measures for simulation Scenario 2 are displayed in Table
	\ref{tab:simulations_gdsc_tab2}. Again, we see that NIG$_{\text{d}}$ performs
	best in terms of MSE. However, NIG$_{\text{d}}^-$ performance is the same 
	here, indicating that the external covariates do not add much here.
	Similarly as in Scenario 1, most of the estimation performance gain over ridge
	is through the top 10\% of the $\beta_{jd}$.

<<simulations_gdsc_tab2>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure for simulation Scenario 2, estimated on the test data (lowest in bold).",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@

<<simulation_gdsc_res3>>=
res <- read.table("results/simulations_gdsc_res3.txt", row.names=NULL)
temp <- res[, 1]
res <- as.matrix(res[, -1])
rownames(res) <- temp
true.chi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.chi <- median(1/res[rownames(res)=="alphad0", 1])
est.lambdad <- apply(res[rownames(res)=="lambdad", c(1, 2)], 2, median)
true.phi <- mean(1/(c(1, 1, 3, 7) %*% t(unname(model.matrix(~ factor(1:4))))))
est.phi <- median(1/res[rownames(res)=="alphaf0", 1])
est.lambdaf <- apply(res[rownames(res)=="lambdaf", c(1, 2)], 2, median)

phi.ratio <- apply(apply(sapply(0:3, function(s) {
  res[rownames(res)==paste0("alphaf" ,s), 2]}) %*%
    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
      res[rownames(res)=="alphaf0", 2]/m}), 2, median)
chi.ratio <- apply(apply(sapply(0:3, function(s) {
  res[rownames(res)==paste0("alphad" ,s), 2]}) %*%
    t(cbind(1, rbind(0, diag(3)))), 2, function(m) {
      res[rownames(res)=="alphad0", 2]/m}), 2, median)
true.ratio <- as.numeric(1/(c(1, 1, 3, 7) %*%
                              t(unname(model.matrix(~ factor(1:4))))))

tab <- cbind(
  apply(
    aggregate(res[substr(rownames(res), 1, 6)=="emsel.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 6)=="emsel.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 6)=="emseh.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 6)=="emseh.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 5)=="emse.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 5)=="emse.", ])),
              FUN="mean")[, -1], 2, median),
  apply(
    aggregate(res[substr(rownames(res), 1, 5)=="pmse.", ], 
              by=list(rownames(res[substr(rownames(res), 1, 5)=="pmse.", ])),
              FUN="mean")[, -1], 2, median))
tab <- round(tab, 3)
tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1] <-
  paste0("\\textbf{", tab[which(tab[, 1]==min(tab[, 1], na.rm=TRUE)), 1], "}")
tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2] <-
  paste0("\\textbf{", tab[which(tab[, 2]==min(tab[, 2], na.rm=TRUE)), 2], "}")
tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3] <-
  paste0("\\textbf{", tab[which(tab[, 3]==min(tab[, 3], na.rm=TRUE)), 3], "}")
tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4] <-
  paste0("\\textbf{", tab[which(tab[, 4]==min(tab[, 4], na.rm=TRUE)), 4], "}")
colnames(tab) <- c("$\\text{EMSE}_{\\text{bottom}}$",
                   "$\\text{EMSE}_{\\text{top}}$",
                   "$\\text{EMSE}_{\\text{total}}$",
                   "$\\text{PMSE}$")
rownames(tab)[c(1, 2)] <- c("NIG$_{\\text{f}+\\text{d}}^-$",
                            "NIG$_{\\text{f}+\\text{d}}$")
@

  Median $\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ are
	are over- and underestimated with \Sexpr{round(est.lambdaf[2], 2)}
	and \Sexpr{round(est.lambdad[2], 2)}, respectively, compared to their true
	value 1.

  NIG$_{\text{f}+\text{d}}^-$ also over- and
	underestimates 
	$\lambda_{\text{feat}}$ and $\lambda_{\text{drug}}$ with median estimates
	\Sexpr{round(est.lambdaf[1], 2)}
	and \Sexpr{round(est.lambdad[1], 2)}, respectively.

  Table 
	\ref{tab:simulations_gdsc_tab3}, with median performance measures, shows
	that NIG$_{\text{f}+\text{d}}$ outperforms the other models in terms of
	prediction and estimation MSE, with again most of the edge over ridge in the
	top 10\% of the $\beta_{jd}$.

<<simulations_gdsc_tab3>>=
library(knitr, quietly=TRUE, verbose=FALSE)
suppressWarnings(library(kableExtra))
options(knitr.kable.NA="", digits=3)
kable(tab, col.names=colnames(tab), align="r",
      caption="Median performance measure for simulation Scenario 3, estimated on the test data (lowest in bold).",
      format="latex", booktabs=TRUE, escape=FALSE) %>%
  kable_styling(latex_options=c("HOLD_position"))
@
  \bibliographystyle{author_short3}  
	\bibliography{refs}
